---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pjrt

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
![R-CMD-check](https://github.com/r-xla/pjrt/actions/workflows/R-CMD-check.yaml/badge.svg)
[![CRAN status](https://www.r-pkg.org/badges/version/pjrt)](https://CRAN.R-project.org/package=pjrt)
[![codecov](https://codecov.io/gh/r-xla/pjrt/branch/main/graph/badge.svg)](https://codecov.io/gh/r-xla/pjrt)
<!-- badges: end -->

The {pjrt} package provides an R interface to [PJRT](https://github.com/openxla/pjrt) ("Pretty much Just another RunTime"), which allows you to run XLA or StableHLO programs on a variety of hardware backends.
For a low-level interface to write StableHLO programs, you can use the [stablehlo](https://github.com/r-xla/stablehlo) package.

## Installation

You can install the development version of {pjrt} from [GitHub](https://github.com/r-xla/pjrt) with:

``` r
# install.packages("pak")
pak::pak("r-xla/pjrt")
```

## Example

Below, we create a simple program in the StableHLO MLIR dialect that adds two single-precision floating point tensors of shape `(2, 2)` and returns the result.

```{r example}
library(pjrt)
src <- r"(
func.func @main(
  %x: tensor<2x2xf32>,
  %y: tensor<2x2xf32>
) -> tensor<2x2xf32> {
  %0 = "stablehlo.add"(%x, %y) : (tensor<2x2xf32>, tensor<2x2xf32>) -> tensor<2x2xf32>
  "func.return"(%0): (tensor<2x2xf32>) -> ()
}
)"
```

The {pjrt} package allows you to create a PJRT program from this string:

```{r}
program <- pjrt_program(src, format = "mlir")
program
```

This program can now be compiled using a client that wraps a plugin, which is a dynamic library implementing the PJRT API.
The plugin is hardware-specific; different hardware backends provide their own plugins.
If not specified explicitly, the default client (CPU plugin if not configured otherwise) is used.

```{r}
program_compiled <- pjrt_compile(program)
```

To run this program, we need to create PJRT buffers, which represent the data (typically tensors).

To create such a tensor, use the `pjrt_buffer()` function.
Because the program above expects `2x2` tensors with element type `f32`, we need to create buffers with this type.

```{r}
x <- matrix(as.double(1:4), nrow = 2)
y <- matrix(as.double(5:8), nrow = 2)
x_buffer <- pjrt_buffer(x, type = "f32")
y_buffer <- pjrt_buffer(y, type = "f32")
```

Having created these buffers, we can now run the program:

```{r}
result <- pjrt_execute(program_compiled, x_buffer, y_buffer)
```

We can convert this result back to an R array:

```{r}
as_array(result)
```

## Platform Support

- **Linux**
  - :white_check_mark: CPU backend is fully supported.
  - :white_check_mark: CUDA (NVIDIA GPU) backend is fully supported.
- **Windows**
  - :warning: Currently only supported via Windows Subsystem for Linux (WSL2).
- **macOS**
  - :white_check_mark: CPU backend is supported.
  - :warning: Metal (Apple GPU) backend is available but not fully functional.

## Acknowledgements

The design of the {pjrt} package was inspired by the [gopjrt](https://github.com/gomlx/gopjrt) implementation.
