---
output: github_document
---

# pjrt

Package website: [release](https://r-xla.github.io/pjrt/) |
[dev](https://r-xla.github.io/pjrt/dev/)

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
pkgload::load_all(quiet = TRUE)
cuda_pkg <- cuda_r_package()
```

<!-- badges: start -->
![R-CMD-check](https://github.com/r-xla/pjrt/actions/workflows/R-CMD-check.yaml/badge.svg)
[![CRAN status](https://www.r-pkg.org/badges/version/pjrt)](https://CRAN.R-project.org/package=pjrt)
[![codecov](https://codecov.io/gh/r-xla/pjrt/branch/main/graph/badge.svg)](https://codecov.io/gh/r-xla/pjrt)
[![r-universe](https://r-xla.r-universe.dev/badges/pjrt)](https://r-xla.r-universe.dev/pjrt)
<!-- badges: end -->

The {pjrt} package provides an R interface to [PJRT](https://github.com/openxla/pjrt) (Pluggable Jit RunTime), which allows you to *run* [XLA](https://openxla.org/) and [stableHLO](https://openxla.org/stablehlo) programs on various hardware backends.
These programs are framework and hardware agnostic, which means they can be generated by ML frameworks such as jax, and run by PJRT on a specified backend (CPU, GPU, etc.).
For a low-level R interface to *create* stableHLO programs, see the [stablehlo](https://github.com/r-xla/stablehlo) package.

## Installation

From r-universe:

```{r, eval = FALSE}
install.packages("pjrt", repos = c("https://r-xla.r-universe.dev", getOption("repos")))
```

### Plugin download

`pjrt` relies on a backend-specific PJRT *plugin* (a shared library) that is not bundled with the package.
The first time you create a client or compile a program for a platform, the matching plugin is downloaded and cached in `tools::R_user_dir("pjrt", "cache")`.

In an interactive session you are asked for confirmation before the download.
The `PJRT_INSTALL` environment variable controls this:

* `PJRT_INSTALL=1`: always download without asking (e.g. in CI, scripts, or Docker builds).
* `PJRT_INSTALL=0`: never download; an error with instructions is raised instead.

In a non-interactive session the plugin is **not** downloaded automatically unless `PJRT_INSTALL=1` is set.
Alternatively, set `PJRT_PLUGIN_PATH_<PLATFORM>` (e.g. `PJRT_PLUGIN_PATH_CPU`) to a local plugin file to skip the download entirely.
See `?pjrt` for the full list of environment variables.

### CUDA

To use the CUDA backend, install the  {`r cuda_pkg`} R package which provides the required CUDA runtime libraries and you only need to have a compatible CUDA driver.

```{r, echo = FALSE, results = "asis"}
cat(sprintf('```r\npak::pak("mlverse/cudatoolkit/%s")\n```', cuda_pkg))
```

Alternatively, install from [r-universe](https://mlverse.r-universe.dev/).

```{r, echo = FALSE, results = "asis"}
cat(sprintf('```r\ninstall.packages("%s", repos = "https://mlverse.r-universe.dev")\n```', cuda_pkg))
```

When the {`r cuda_pkg`} package is not installed, the correct runtime libraries need to be installed on the system, which can be difficult to set up.
The specific versions of the CUDA runtime libraries provided with {`r cuda_pkg`} are provided [here](`r sprintf("https://github.com/mlverse/cudatoolkit/blob/main/%s/inst/components.tsv", cuda_pkg)`).

**Troubleshooting**

To trouble-shoot the CUDA installation, run the following in a new R session for maximum debug output.

```{r, eval = FALSE}
Sys.setenv(PJRT_DEBUG = "1", TF_CPP_MIN_LOG_LEVEL = "0")
pjrt::pjrt_buffer(1, device = "cuda")
```

Note that if another package is using a different cudatoolkit package (e.g. when using {torch}), there might be some issues, so in this case it's best to run separate R processes.


## Quick Start

Below, we create and run a stableHLO program that adds two `f32` tensors of shape `2x2`.

```{r example}
library(pjrt)
src <- r"(
func.func @main(
  %x: tensor<2x2xf32>,
  %y: tensor<2x2xf32>
) -> tensor<2x2xf32> {
  %0 = "stablehlo.add"(%x, %y) : (tensor<2x2xf32>, tensor<2x2xf32>) -> tensor<2x2xf32>
  "func.return"(%0): (tensor<2x2xf32>) -> ()
}
)"
program <- pjrt_program(src, format = "mlir")
program
executable <- pjrt_compile(program, device = "cpu")

x <- pjrt_buffer(c(1, 2, 3, 4), shape = c(2, 2), dtype = "f32")
x
y <- pjrt_buffer(c(5, 6, 7, 8), shape = c(2, 2), dtype = "f32")
y

pjrt_execute(executable, x, y)
```

## Main Features

* Compile stableHLO programs into hardware-specific executables.
* Provide a runtime to execute compiled programs.
* Convert buffers to and from R arrays and vectors.
* Read and write buffers using the [safetensors](https://github.com/mlverse/safetensors) format.
* Operations are executed asynchronously.

## Platform Support

* **Linux (x86_64)**
  * :white_check_mark: CPU backend is fully supported.
  * :white_check_mark: CUDA (NVIDIA GPU) backend is fully supported.
* **Linux (ARM)**
  * :white_check_mark: CPU backend is fully supported.
  * :x: GPU is not supported.
* **Windows**
  * :white_check_mark: CPU backend is fully supported.
  * :warning: GPU is only supported via Windows Subsystem for Linux (WSL2).
* **macOS**
  * :white_check_mark: CPU backend is supported.
  * :warning: Metal (Apple GPU) backend is available but not fully functional.

## Acknowledgements

* The development of this package is supported by [MaRDI](https://www.mardi4nfdi.de/about/mission).
* Without [OpenXLA](https://openxla.org/), none of this would be possible.
* The design of the {pjrt} package was inspired by the [gopjrt](https://github.com/gomlx/gopjrt) implementation.
* The project also uses various components from OpenXLA:
  * [PJRT C API](https://github.com/openxla/xla/blob/main/xla/pjrt/c/pjrt_c_api.h).
  * [PJRT C API FFI Extension](https://github.com/openxla/xla/blob/main/xla/pjrt/c/pjrt_c_api_ffi_extension.h).
  * Various protobuf files, see `./tools/copy-proto.R` for which ones.
  * Plugin implementations for CPU and CUDA (we are using the builds from [zml/pjrt-artifacts](https://github.com/zml/pjrt-artifacts/)).
* For Metal, we are using the plugin implementation from [jax-metal](https://pypi.org/project/jax-metal/).
