---
output: github_document
---

# pjrt

Package website: [release](https://r-xla.github.io/pjrt/) |
[dev](https://r-xla.github.io/pjrt/dev/)

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
pkgload::load_all(quiet = TRUE)
```

<!-- badges: start -->
![R-CMD-check](https://github.com/r-xla/pjrt/actions/workflows/R-CMD-check.yaml/badge.svg)
[![CRAN status](https://www.r-pkg.org/badges/version/pjrt)](https://CRAN.R-project.org/package=pjrt)
[![codecov](https://codecov.io/gh/r-xla/pjrt/branch/main/graph/badge.svg)](https://codecov.io/gh/r-xla/pjrt)
[![r-universe](https://r-xla.r-universe.dev/badges/pjrt)](https://r-xla.r-universe.dev/pjrt)
<!-- badges: end -->

{pjrt} is an R interface to [PJRT](https://openxla.org/xla/pjrt), the runtime API of [OpenXLA](https://openxla.org/).
It compiles [StableHLO](https://openxla.org/stablehlo) (or HLO) programs with the XLA compiler into executables for a specific device and runs them there.
StableHLO programs do not depend on the framework that produced them, so a program exported from, e.g., JAX can be run by {pjrt}.
To *create* StableHLO programs from R, see the [stablehlo](https://github.com/r-xla/stablehlo) package.

{pjrt} is a low-level runtime.
For array computing in R with JIT compilation and automatic differentiation, use [anvl](https://github.com/r-xla/anvl), which builds on {pjrt}.

## Installation

```{r, eval = FALSE}
install.packages("pjrt", repos = c("https://r-xla.r-universe.dev", getOption("repos")))
pjrt::install_pjrt()
```

Each backend (CPU or CUDA) is a PJRT *plugin*: a shared library that is not bundled with the package but downloaded separately.
`install_pjrt()` downloads the CPU plugin and, if an NVIDIA GPU is detected, the CUDA plugin together with the {`r cuda_r_package()`} package, which provides the CUDA runtime libraries.
Without `install_pjrt()`, a plugin is downloaded the first time its platform is used.
See `?pjrt` for the environment variables that control this.

## Quick Start

Below, we compile and run a StableHLO program that adds two `f32` tensors of shape `(2x2)`.

```{r example}
library(pjrt)
src <- r"(
func.func @main(
  %x: tensor<2x2xf32>,
  %y: tensor<2x2xf32>
) -> tensor<2x2xf32> {
  %0 = "stablehlo.add"(%x, %y) : (tensor<2x2xf32>, tensor<2x2xf32>) -> tensor<2x2xf32>
  "func.return"(%0): (tensor<2x2xf32>) -> ()
}
)"
program <- pjrt_program(src, format = "mlir")
program
executable <- pjrt_compile(program, device = "cpu")

x <- pjrt_buffer(c(1, 2, 3, 4), shape = c(2, 2), dtype = "f32")
x
y <- pjrt_buffer(c(5, 6, 7, 8), shape = c(2, 2), dtype = "f32")
y

pjrt_execute(executable, x, y)
```

## Main Features

* Compile StableHLO and HLO programs into device-specific executables.
* Execute them asynchronously on CPU or CUDA devices.
* Convert between R arrays and device buffers, with range checks and `NA` handling.
* Register custom XLA FFI handlers.
* Read and write buffers using the [safetensors](https://github.com/mlverse/safetensors) format.

## Platform Support

```{r, echo = FALSE}
support <- platform_support()
backends <- c(cpu = "CPU", cuda = "CUDA")
symbols <- c(yes = "✓", no = "✗", wsl2 = "WSL2")
support[names(backends)] <- lapply(support[names(backends)], function(x) unname(symbols[x]))
names(support) <- c("OS", "Architecture", backends)
knitr::kable(support, align = "llcc")
```

✓ supported &nbsp;·&nbsp; WSL2: via the Windows Subsystem for Linux &nbsp;·&nbsp; ✗ not supported

## Acknowledgements

* The development of this package is supported by [MaRDI](https://www.mardi4nfdi.de/about/mission).
* Without [OpenXLA](https://openxla.org/), none of this would be possible.
* The design of the {pjrt} package was inspired by the [gopjrt](https://github.com/gomlx/gopjrt) implementation.
* The project also uses various components from OpenXLA:
  * [PJRT C API](https://github.com/openxla/xla/blob/main/xla/pjrt/c/pjrt_c_api.h).
  * [PJRT C API FFI Extension](https://github.com/openxla/xla/blob/main/xla/pjrt/c/pjrt_c_api_ffi_extension.h).
  * Various protobuf files, see `./tools/copy-proto.R` for which ones.
  * Plugin implementations for CPU and CUDA (we are using the builds from [zml/pjrt-artifacts](https://github.com/zml/pjrt-artifacts/), and our own [Windows build](https://github.com/r-xla/pjrt-builds)).
