---
title: "Running a JAX Model in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running a JAX Model in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = pjrt::plugins_downloaded("cpu") && pjrt::stablehlo_opt_available()
)
```

StableHLO is a portable representation of a computation, so a model does not
have to be trained in the same language it is deployed in.
In this article we export a small MLP from JAX and run it in R with {pjrt}.

The interesting part is *shape polymorphism*: JAX can export a model whose
batch dimension is left symbolic, so a single artifact serves any batch size.
XLA, on the other hand, compiles for static shapes.
Bridging the two is what `pjrt_refine_shapes()` does.

```{r}
library(pjrt)
```

## Exporting from JAX

On the Python side we define an MLP and export it with a symbolic batch
dimension `batch`.
Everything except the leading dimension of `x` is static.

```python
import jax
import jax.numpy as jnp
from jax import export

def mlp(w1, b1, w2, b2, x):
    h = jnp.tanh(x @ w1 + b1)
    return h @ w2 + b2

(batch,) = export.symbolic_shape("batch")

exported = export.export(jax.jit(mlp))(
    jax.ShapeDtypeStruct((3, 4), jnp.float32),      # w1
    jax.ShapeDtypeStruct((4,), jnp.float32),        # b1
    jax.ShapeDtypeStruct((4, 2), jnp.float32),      # w2
    jax.ShapeDtypeStruct((2,), jnp.float32),        # b2
    jax.ShapeDtypeStruct((batch, 3), jnp.float32),  # x
)

with open("jax-mlp-dynamic.mlir", "w") as f:
    f.write(exported.mlir_module())
```

The resulting module ships with {pjrt} so that this article can be run without
a Python installation.

```{r}
path <- system.file("programs/jax-mlp-dynamic.mlir", package = "pjrt")
program <- pjrt_program(path = path)
```

Its `main` function takes the four parameter tensors plus `x`, whose first
dimension is `?`, i.e. not known yet.
The symbolic dimension is threaded through the program as an `i32` value and
guarded by a `stablehlo.custom_call @shape_assertion`.

```{r}
print(program, n = 8)
```

Handing this program to the compiler does not work -- XLA needs to know how
much memory every intermediate tensor requires.

```{r, error=TRUE}
pjrt_compile(program)
```

## Concretizing the shapes

`pjrt_refine_shapes()` rewrites the argument types of `main` and propagates
them through the program, i.e. it fills in every `?` that follows from the new
argument types.
The types are given in MLIR syntax, one per argument of `main`.

```{r}
batch_size <- 5

refined <- pjrt_refine_shapes(
  program,
  c(
    "tensor<3x4xf32>",
    "tensor<4xf32>",
    "tensor<4x2xf32>",
    "tensor<2xf32>",
    sprintf("tensor<%dx3xf32>", batch_size)
  )
)

print(refined, n = 8)
```

The dynamic ops are gone: `stablehlo.dynamic_broadcast_in_dim` became a plain
`broadcast_in_dim`, the shape assertion was checked and erased, and every
tensor type is static.
This program compiles.

```{r}
executable <- pjrt_compile(refined)
```

## Running it

The weights and the input are ordinary R arrays.
Note that {pjrt} takes care of the row-major/column-major conversion.

```{r}
w1 <- pjrt_buffer(matrix(seq(0.1, 1.2, length.out = 12), nrow = 3), dtype = "f32")
b1 <- pjrt_buffer(rep(0, 4), dtype = "f32")
w2 <- pjrt_buffer(matrix(seq(-0.4, 0.4, length.out = 8), nrow = 4), dtype = "f32")
b2 <- pjrt_buffer(c(1, -1), dtype = "f32")
x <- pjrt_buffer(matrix(seq(0, 1, length.out = batch_size * 3), nrow = batch_size), dtype = "f32")

as_array(pjrt_execute(executable, w1, b1, w2, b2, x))
```

## One export, many shapes

Because the refinement happens on the StableHLO program rather than on the
export, the same artifact can be specialized for a different batch size
whenever one is needed.

```{r}
predict_batch <- function(x) {
  refined <- pjrt_refine_shapes(
    program,
    c(
      "tensor<3x4xf32>",
      "tensor<4xf32>",
      "tensor<4x2xf32>",
      "tensor<2xf32>",
      sprintf("tensor<%dx3xf32>", nrow(x))
    )
  )
  executable <- pjrt_compile(refined)
  as_array(pjrt_execute(
    executable,
    w1,
    b1,
    w2,
    b2,
    pjrt_buffer(x, dtype = "f32")
  ))
}

predict_batch(matrix(0.5, nrow = 2, ncol = 3))
predict_batch(matrix(0.5, nrow = 7, ncol = 3))
```

Compilation is expensive, so in practice you would cache the executable per
batch size instead of recompiling on every call.

## Other transformations

Shape refinement is one of the passes provided by StableHLO's `stablehlo-opt`
tool, which {pjrt} downloads on demand.
`pjrt_optimize()` gives access to all of them, e.g. to constant folding:

```{r}
pjrt_optimize(
  "
  func.func @main() -> tensor<2xf32> {
    %0 = stablehlo.constant dense<[1.0, 2.0]> : tensor<2xf32>
    %1 = stablehlo.constant dense<[3.0, 4.0]> : tensor<2xf32>
    %2 = stablehlo.add %0, %1 : tensor<2xf32>
    return %2 : tensor<2xf32>
  }
  ",
  passes = "stablehlo-aggressive-folder"
)
```

The available passes are:

```{r}
stablehlo_opt_passes()
```
